---
title: "ASD Phenome Scan"
author: "Louis Klein"
date: "10/11/2020"
output: html_notebook
css: table-style.css
---

setup document and load files

```{r settings, message=FALSE, warning=FALSE}
library(easypackages)
suppressPackageStartupMessages(
  libraries("knitr","tidyverse","magrittr")
)

knitr::opts_chunk$set(
  autodep = TRUE, # analyse chunk dependencies by automatically detecting global variables
  cache = TRUE, # cache code chunks so that recompute will only occur on change
  cache.comments = FALSE, # changing comments will not invalidate the cache
  message = FALSE, # discard messages
  warning = FALSE # discard warnings
  )

options(max.print = 3000)

### Load data ------------------------------------------------------------------

load(file = "datafiles/workspace_10-11-20.RData")
glimpse(phenomes)
```

import data and replace negative values with `NA`s

```{r}
# read data into R and convert missing to NAs
phenomes <- read_csv('datafiles/AGRE_clinical_data20201016.csv',
                     col_names = TRUE,
                     col_types = cols(
                       individualid = col_factor()
                     )) %>%
  mutate(
    across(
      where(is.numeric),
    ~ replace(.,.<0,NA)
  ))

# spec(phenomes)
```

remove non-target variables, replace RA codes with `NA`s, and collapse Likert-style variables

```{r}
# drop unusable columns and prepare dataset
phenomes_clean <- phenomes %>%
  select(-c(frequency_of_seizures,hospitalization_year,hospitalization_diagnosis,
            other_psychiatric_problems,other_psychiatric_problems_typ,cur_supp_and_alter_comm,
            special_diet_prev_cmmt,gluten_free_diet_duration_commen,other_diet_therapies,
            other_diet_preferences,depression_age_at_diagnosis,depressive_symp_age_onset,
            depression_age_at_diagnosis,bmd_age_of_onset,bmd_age_at_diagnosis,anxiety_age_of_onset,
            anxiety_age_at_diagnosis,age_of_seizure_onset_years,ocd_age_of_onset,
            ocd_age_at_diagnosis,adhd_age_of_onset,adhd_age_at_diagnosis,behavioral_problem_age_of_onset,
            behavioral_problem_age_at_diagno,sleep_disorder_age_of_onset,sleep_disorder_age_at_diagnosis,
            sleep_study,roll_months,sit_months,walk_months,first_words_months,first_phrases_months,
            toliet_training_day_bladder_mont,toliet_training_day_bowel_months,special_diet_current,
            toliet_training_night_bladder_mo,toliet_training_night_bowel_mont,number_of_seizures)) %>%
  mutate_at(
    vars(alcohol_use,previous_alcohol_use,tobacco_use,previous_tobacco_use,drug_use,
         previous_drug_use,imaging_study_3),
  ~ case_when(.>0~1,TRUE~.)) %>%
  mutate_all(~str_replace_all(.,"9",NA_character_)) %>%
  mutate_at(vars(-("individualid")),~as.numeric(.)) %>%
  mutate(individualid = as.factor(individualid))

glimpse(phenomes_clean)
```

examine missingness by variable, n.b. highly sparse data structure

```{r}
pmiss <- unlist(lapply(phenomes[,2:122], function(p) sum(is.na(p))))/nrow(phenomes[,2:122])
sort(pmiss[pmiss > 0], decreasing = TRUE)
```

apply k-means clustering to dichotomise target continuous variables

```{r}
library(tidymodels)

k_roll <- phenomes %>%
  column_to_rownames(var="individualid") %>%
  select(roll_months) %>%
  drop_na() %>%
  kmeans(.,2)

k_sit <- phenomes %>%
  column_to_rownames(var="individualid") %>%
  select(sit_months) %>%
  drop_na() %>%
  kmeans(.,2)

k_walk <- phenomes %>%
  column_to_rownames(var="individualid") %>%
  select(walk_months) %>%
  drop_na() %>%
  kmeans(.,2)

k_first_words <- phenomes %>%
  column_to_rownames(var="individualid") %>%
  select(first_words_months) %>%
  drop_na() %>%
  kmeans(.,2)

k_first_phrases <- phenomes %>%
  column_to_rownames(var="individualid") %>%
  select(first_phrases_months) %>%
  drop_na() %>%
  kmeans(.,2)

k_tt_day_bladder <- phenomes %>%
  column_to_rownames(var="individualid") %>%
  select(toliet_training_day_bladder_mont) %>%
  drop_na() %>%
  kmeans(.,2)

k_tt_day_bowel <- phenomes %>%
  column_to_rownames(var="individualid") %>%
  select(toliet_training_day_bowel_months) %>%
  drop_na() %>%
  kmeans(.,2)

k_tt_night_bladder <- phenomes %>%
  column_to_rownames(var="individualid") %>%
  select(toliet_training_night_bladder_mo) %>%
  drop_na() %>%
  kmeans(.,2)

k_tt_night_bowel <- phenomes %>%
  column_to_rownames(var="individualid") %>%
  select(toliet_training_night_bowel_mont) %>%
  drop_na() %>%
  kmeans(.,2)

# Extract and tidy clusters -----------------------------------------------

k_roll_dichot <- k_roll$cluster %>% 
  enframe() %>%
  rename(id = "name",
         roll = "value")

k_sit_dichot <- k_sit$cluster %>% 
  enframe() %>%
  rename(id = "name",
         sit = "value")

k_walk_dichot <- k_walk$cluster %>% 
  enframe() %>%
  rename(id = "name",
         walk = "value")

k_first_words_dichot <- k_first_words$cluster %>% 
  enframe() %>%
  rename(id = "name",
         words = "value")

k_first_phrases_dichot <- k_first_phrases$cluster %>% 
  enframe() %>%
  rename(id = "name",
         phrases = "value")

k_tt_day_bladder_dichot <- k_tt_day_bladder$cluster %>% 
  enframe() %>%
  rename(id = "name",
         diem_bladder = "value")

k_tt_day_bowel_dichot <- k_tt_day_bowel$cluster %>% 
  enframe() %>%
  rename(id = "name",
         diem_bowel = "value")

k_tt_night_bladder_dichot <- k_tt_night_bladder$cluster %>% 
  enframe() %>%
  rename(id = "name",
         nocte_bladder = "value")

k_tt_night_bowel_dichot <- k_tt_night_bowel$cluster %>% 
  enframe() %>%
  rename(id = "name",
         nocte_bowel = "value")

# Join clusters into dataframe --------------------------------------------

k_dichotomisation <- k_roll_dichot %>%
  full_join(k_sit_dichot, by = "id") %>%
  full_join(k_walk_dichot, by = "id") %>%
  full_join(k_first_words_dichot, by = "id") %>%
  full_join(k_first_phrases_dichot, by = "id") %>%
  full_join(k_tt_day_bladder_dichot, by = "id") %>%
  full_join(k_tt_day_bowel_dichot, by = "id") %>%
  full_join(k_tt_night_bladder_dichot, by = "id") %>%
  full_join(k_tt_night_bowel_dichot, by = "id")
```

recode dichotomised data to reflect developmental delay

```{r}
k_dichotomisation_recoded <- k_dichotomisation %>%
  mutate_at(
    vars(-("id")),
  ~ case_when(.==2~"delay",
              TRUE~NA_character_
            ))
```

manually recode and rename variables in preparation for topic modelling

```{r}
phenomes_prep <- phenomes_clean %>%
  mutate(
    id = individualid,
    special_diet_prev = case_when(special_diet_prev==1~"past-diet",TRUE~NA_character_),
    casein_free_diet = case_when(casein_free_diet==1~"casein-free",TRUE~NA_character_),
    gluten_free_diet = case_when(gluten_free_diet==1~"gluten-free",TRUE~NA_character_),
    dairy_free_diet = case_when(dairy_free_diet==1~"dairy-free",TRUE~NA_character_),
    egg_free_diet = case_when(egg_free_diet==1~"egg-free",TRUE~NA_character_),
    acoustic = case_when(acoustic>=3~"high-acoustic",acoustic<=2~"low-acoustic",TRUE~NA_character_),
    tactile = case_when(tactile>=3~"high-tactile",tactile<=2~"low-tactile",TRUE~NA_character_),
    pain = case_when(pain>=3~"high-pain",pain<=2~"low-pain",TRUE~NA_character_),
    visual = case_when(visual>=3~"high-visual",visual<=2~"low-visual",TRUE~NA_character_),
    floppy_infant = case_when(floppy_infant==1~"infant-floppy",TRUE~NA_character_),
    stiff_infant = case_when(stiff_infant==1~"infant-stiff",TRUE~NA_character_),
    irritable_unconsolable_infant = case_when(irritable_unconsolable_infant==1~"infant-irritable",TRUE~NA_character_),
    lethargic_overly_sleepy_infant = case_when(lethargic_overly_sleepy_infant==1~"infant-lethargic",TRUE~NA_character_),
    craniofacial_problems = case_when(craniofacial_problems==1~"abnorm-craniofacial",TRUE~NA_character_),
    skin_abnormalities = case_when(skin_abnormalities==1~"abnorm-skin",TRUE~NA_character_),
    teeth_abnormalities = case_when(teeth_abnormalities==1~"abnorm-teeth",TRUE~NA_character_),
    eye_abnormalities_structural = case_when(eye_abnormalities_structural==1~"abnorm-eye",TRUE~NA_character_),
    ear_abnormalities_structural = case_when(ear_abnormalities_structural==1~"abnorm-ear",TRUE~NA_character_),
    ear_pe_tubes = case_when(ear_pe_tubes==1~"abnorm-ear-pe",TRUE~NA_character_),
    bone_abnormalities = case_when(bone_abnormalities==1~"abnorm-bone",TRUE~NA_character_),
    joint_abnormalities = case_when(joint_abnormalities==1~"abnorm-joint",TRUE~NA_character_),
    respiratory_problems = case_when(respiratory_problems==1~"abnorm-respiratory",TRUE~NA_character_),
    asthma = case_when(asthma==1~"asthma",TRUE~NA_character_),
    allergies = case_when(allergies==1~"allergies",TRUE~NA_character_),
    heart_problems = case_when(heart_problems==1~"abnorm-heart",TRUE~NA_character_),
    gastrointestinal_problems = case_when(gastrointestinal_problems==1~"abnorm-gastro",TRUE~NA_character_),
    kidney_urinary_problems = case_when(kidney_urinary_problems==1~"abnorm-renal",TRUE~NA_character_),
    genital_problems = case_when(genital_problems==1~"abnorm-genital",TRUE~NA_character_),
    endocrine_problems_growth = case_when(endocrine_problems_growth==1~"endocrine-growth",TRUE~NA_character_),
    endocrine_problems_thyroid = case_when(endocrine_problems_thyroid==1~"endocrine-thyroid",TRUE~NA_character_),
    endocrine_problems_sex_hormone = case_when(endocrine_problems_sex_hormone==1~"endocrine-sex",TRUE~NA_character_),
    endocrine_problems_diabetes = case_when(endocrine_problems_diabetes==1~"endocrine-diabetes",TRUE~NA_character_),
    abnormal_growth_pattern = case_when(abnormal_growth_pattern==1~"abnorm-growth",TRUE~NA_character_),
    vision_problems = case_when(vision_problems==1~"vision",TRUE~NA_character_),
    hearing_problems = case_when(hearing_problems==1~"hearing",TRUE~NA_character_),
    hearing_assessed_by_baer = case_when(hearing_assessed_by_baer==1~"hearing-baer",TRUE~NA_character_),
    hearing_assessed_by_behavioral = case_when(hearing_assessed_by_behavioral==1~"hearing-behavioural",TRUE~NA_character_),
    hearing_assessment_result = case_when(hearing_assessment_result==1~"hearing-assessment",TRUE~NA_character_),
    febrile_seizures = case_when(febrile_seizures==1~"seizures-febrile",TRUE~NA_character_),
    other_seizures = case_when(other_seizures==1~"seizures-other",TRUE~NA_character_),
    intractability_of_seizures = case_when(intractability_of_seizures==1~"seizures-intractable",TRUE~NA_character_),
    handedness = case_when(handedness==1~"handed-right",handedness==2~"handed-left",TRUE~NA_character_),
    cerebral_abnormalities = case_when(cerebral_abnormalities==1~"abnorm-cerebral",TRUE~NA_character_),
    movement_abnormalities = case_when(movement_abnormalities==1~"abnorm-movement",TRUE~NA_character_),
    stereotypies = case_when(stereotypies==1~"stereotypies",TRUE~NA_character_),
    coordination_abnormalities = case_when(coordination_abnormalities==1~"abnorm-coordination",TRUE~NA_character_),
    gait_abnormalities = case_when(gait_abnormalities==1~"abnorm-gait",TRUE~NA_character_),
    sensory_abnormalities = case_when(sensory_abnormalities==1~"abnorm-sensory",TRUE~NA_character_),
    karyotype = case_when(karyotype==1~"test-karyotype",TRUE~NA_character_),
    metabolic = case_when(metabolic==1~"test-metabolic",TRUE~NA_character_),
    other_molecular = case_when(other_molecular==1~"test-other",TRUE~NA_character_),
    imaging_study_3 = case_when(imaging_study_3==1~"test-imaging",TRUE~NA_character_),
    depressive_symptoms = case_when(depressive_symptoms==1~"symptoms-depression",TRUE~NA_character_),
    depression_diagnosis = case_when(depression_diagnosis==1~"diagnosis-depression",TRUE~NA_character_),
    bipolar_mood_disorder_symptoms = case_when(bipolar_mood_disorder_symptoms==1~"symptoms-bpd",TRUE~NA_character_),
    bmd_diagnosis = case_when(bmd_diagnosis==1~"diagnosis-bmd",TRUE~NA_character_),
    anxiety_symptoms = case_when(anxiety_symptoms==1~"symptoms-anxiety",TRUE~NA_character_),
    anxiety_diagnosis = case_when(anxiety_diagnosis==1~"diagnosis-anxiety",TRUE~NA_character_),
    ocd_symptoms = case_when(ocd_symptoms==1~"symptoms-ocd",TRUE~NA_character_),
    ocd_diagnosis = case_when(ocd_diagnosis==1~"diagnosis-ocd",TRUE~NA_character_),
    adhd_symptoms = case_when(adhd_symptoms==1~"symptoms-adhd",TRUE~NA_character_),
    adhd_diagnosis = case_when(adhd_diagnosis==1~"diagnosis-adhd",TRUE~NA_character_),
    behavioral_problem_symptoms = case_when(behavioral_problem_symptoms==1~"symptoms-odd",TRUE~NA_character_),
    behavioral_problem_diagnosis = case_when(behavioral_problem_diagnosis==1~"diagnosis-odd",TRUE~NA_character_),
    sleep_disorder_symptoms = case_when(sleep_disorder_symptoms==1~"symptoms-sleep",TRUE~NA_character_),
    sleep_disorder_diagnosis = case_when(sleep_disorder_diagnosis==1~"diagnosis-sleep",TRUE~NA_character_),
    difficulty_falling_asleep = case_when(difficulty_falling_asleep==1~"sleep-insomnia",TRUE~NA_character_),
    interupted_sleep = case_when(interupted_sleep==1~"sleep-interrupted",TRUE~NA_character_),
    night_terrors = case_when(night_terrors==1~"sleep-terrors",TRUE~NA_character_),
    excessive_movements = case_when(excessive_movements==1~"sleep-fidgeting",TRUE~NA_character_),
    history_of_being_hospitalized = case_when(history_of_being_hospitalized==1~"past-hospital",TRUE~NA_character_),
    dietary_preferences = case_when(dietary_preferences==1~"diet-preferences",TRUE~NA_character_),
    extremely_picky = case_when(extremely_picky==1~"diet-picky",TRUE~NA_character_),
    high_carbohydrate = case_when(high_carbohydrate==1~"diet-carbohydrates",TRUE~NA_character_),
    high_salt = case_when(high_salt==1~"diet-salt",TRUE~NA_character_),
    high_sweet = case_when(high_sweet==1~"diet-sugar",TRUE~NA_character_),
    specific_texture = case_when(specific_texture==1~"diet-texture",TRUE~NA_character_),
    pica_observed = case_when(pica_observed==1~"diet-pica",TRUE~NA_character_),
    alcohol_use = case_when(alcohol_use==1~"current-alcohol",TRUE~NA_character_),
    previous_alcohol_use = case_when(previous_alcohol_use==1~"past-alcohol",TRUE~NA_character_),
    tobacco_use = case_when(tobacco_use==1~"current-tobacco",TRUE~NA_character_),
    previous_tobacco_use = case_when(previous_tobacco_use==1~"past-tobacco",TRUE~NA_character_),
    drug_use = case_when(drug_use==1~"current-drugs",TRUE~NA_character_),
    previous_drug_use = case_when(previous_drug_use==1~"past-drugs",TRUE~NA_character_)
  ) %>%
  full_join(k_dichotomisation_recoded %>%
              mutate(
                roll_months = case_when(roll=="delay"~"delay-roll",TRUE~NA_character_),
                sit_months = case_when(sit=="delay"~"delay-sit",TRUE~NA_character_),
                walk_months = case_when(walk=="delay"~"delay-walk",TRUE~NA_character_),
                first_words_months = case_when(words=="delay"~"delay-words",TRUE~NA_character_),
                first_phrases_months = case_when(phrases=="delay"~"delay-phrases",TRUE~NA_character_),
                toilet_training_day_bladder = case_when(diem_bladder=="delay"~"delay-bladder-diem",TRUE~NA_character_),
                toilet_training_day_bowel = case_when(diem_bowel=="delay"~"delay-bowel-diem",TRUE~NA_character_),
                toilet_training_night_bladder = case_when(nocte_bladder=="delay"~"delay-bladder-nocte",TRUE~NA_character_),
                toilet_training_night_bowel = case_when(nocte_bowel=="delay"~"delay-bowel-nocte",TRUE~NA_character_),
              ) %>%
              select(id,roll_months,sit_months,walk_months,first_words_months,first_phrases_months,
                     toilet_training_day_bladder,toilet_training_day_bowel,toilet_training_night_bladder,
                     toilet_training_night_bowel),
            by="id"
            ) %>%
  select(!individualid) %>%
  drop_na(id)

glimpse(phenomes_prep)
```

make pseudo EMRs and prepare components for topic modelling

```{r}
library(quanteda)
library(tidytext)

phenomes_emr <- phenomes_prep %>%
  column_to_rownames(var="id") %>%
  unite(.,col = "emr", 1:93,na.rm=TRUE, sep = " ") %>%
  filter(!rapportools::is.empty(emr)==TRUE) %>%
  rownames_to_column(var = "id")

# using the quanteda package because tidytext splits hyphenation by default
phenomes_corpus <- corpus(phenomes_emr$emr,docnames=phenomes_emr$id)
phenomes_ngrams <- tokens_ngrams(tokens(phenomes_corpus),n=1,concatenator=" ")
phenomes_dfm <- dfm(phenomes_ngrams)

# convert corpus into sparse matrix
phenomes_sparse <- phenomes_dfm %>% 
  tidytext::tidy() %>%
  cast_sparse(document,term,count)

head(phenomes_emr$emr)
```

test a range of numbers of clusters using a parallel process

```{r}
library(stm)
library(furrr)

plan(multiprocess)

phenomes_models <- data_frame(K = c(6,8,10,12,14,16,18,20,22,24)) %>%
  mutate(topic_model = future_map(K, ~ stm(phenomes_sparse,
                                           K=.,verbose=FALSE)
                                  ))

head(phenomes_models)
```

calculate statistics for investigating the appropriate number of topics to model

```{r}
heldout <- make.heldout(phenomes_sparse)

k_result <- phenomes_models %>%
  mutate(exclusivity = map(topic_model, exclusivity),
         semantic_coherence = map(topic_model, semanticCoherence, phenomes_sparse),
         eval_heldout = map(topic_model, eval.heldout, heldout$missing),
         residual = map(topic_model, checkResiduals, phenomes_sparse),
         bound =  map_dbl(topic_model, function(x) max(x$convergence$bound)),
         lfact = map_dbl(topic_model, function(x) lfactorial(x$settings$dim$K)),
         lbound = bound + lfact,
         iterations = map_dbl(topic_model, function(x) length(x$convergence$bound)))

k_result
```

examine coherence of topic models

```{r}
k_result %>%
  transmute(K,
            `Lower bound` = lbound,
            Residuals = map_dbl(residual, "dispersion"),
            `Semantic coherence` = map_dbl(semantic_coherence, mean),
            `Held-out likelihood` = map_dbl(eval_heldout, "expected.heldout")) %>%
  gather(Metric, Value, -K) %>%
  ggplot(aes(K, Value, color = Metric)) +
  geom_line(size = 1.5, alpha = 0.7, show.legend = FALSE) +
  facet_wrap(~Metric, scales = "free_y") +
  labs(x = "K (number of topics)",
       y = NULL,
       title = "Model diagnostics by number of topics",
       subtitle = "These diagnostics indicate that 18 is an appropriate number of topics")
```

the held-out likelihood is high and stable from 14 to 24 topics however it spikes at 18
the held-out likelihood is coincident with a minimum for the residuals, suggesting that the variance is being optimsed at 18 topics
semantic coherence stabilises for 16-18 topic range

compare feature exclusivity versus semantic coherence

```{r}
k_result %>%
  select(K, exclusivity, semantic_coherence) %>%
  filter(K %in% c(14,16,18)) %>%
  unnest(cols = c(exclusivity, semantic_coherence)) %>%
  mutate(K = as.factor(K)) %>%
  ggplot(aes(semantic_coherence, exclusivity, color = K)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(x = "Semantic coherence",
       y = "Exclusivity",
       title = "Comparing exclusivity and semantic coherence",
       subtitle = "Models with fewer topics have higher semantic coherence for more topics, but lower exclusivity")
```

no real trade-off between exclusivity and coherence

extracting the model for number of topics = 18

```{r}
phenomes_tm <- k_result %>% 
  filter(K == 18) %>% 
  pull(topic_model) %>% 
  .[[1]]

phenomes_beta <- phenomes_tm %>%
  tidytext::tidy()

phenomes_gamma <- phenomes_tm %>%
  tidy(matrix="gamma",document_names=rownames(phenomes_sparse))
```

explore model

```{r}
beta_terms <- phenomes_beta %>%
  arrange(beta) %>%
  group_by(topic) %>%
  top_n(12, beta) %>%
  arrange(-beta) %>%
  select(topic, term) %>%
  summarise(terms = list(term)) %>%
  mutate(terms = map(terms, paste, collapse = ", ")) %>% 
  unnest(cols = c(terms))

gamma_terms <- phenomes_gamma %>%
  group_by(topic) %>%
  summarise(gamma = mean(gamma)) %>%
  arrange(desc(gamma)) %>%
  left_join(beta_terms, by = "topic") %>%
  mutate(topic = paste0("Topic ", topic),
         topic = reorder(topic, gamma))
```

build table of results

```{r}
library(kableExtra)

gamma_terms %>%
  select(topic, gamma, terms) %>%
  kable(format = "html",digits = 3, 
        col.names = c("Topic", "Expected topic proportion", "Top 7 terms"),
        caption = "Dummy text 1",
        align = c("l","c","l"),
        ) %>%
  kable_styling(full_width = TRUE,bootstrap_options = c("condensed","responsive")) %>%
  footnote(general_title = "Note.",general = "Dummy text 2",footnote_as_chunk = TRUE)
```

```{r}
phenomes_tidy %>%
    count(document, term, sort = TRUE) %>%
    bind_tf_idf(term, document, n) %>%
    arrange(-tf_idf) %>%
    group_by(document) %>%
    top_n(10) %>%
    ungroup
```

run latent dirichlet allocation process more robust model params for k=18 and write outputs

```{r}
library(tm)
library(topicmodels)

# generate the LDA model
phenomes_lda <- LDA(phenomes_dfm,k=18,method="Gibbs",
                    control=list(nstart=5,seed=list(254672,109,122887,145629037,2), 
                                 best=TRUE,burnin=1000,iter=2000,thin=500))

phenomes_terms <- as.matrix(terms(phenomes_lda,16))
phenomes_topics <- as.matrix(topics(phenomes_lda))

# make primary output table
phenomes_probs <- as.data.frame(phenomes_lda@gamma)
rownames(phenomes_probs) <- rownames(phenomes_sparse)

# write outputs to .csv
write.csv(phenomes_topics,file=paste0("lda_gibbs_",18,"_docs-to-topics.csv"))
write.csv(phenomes_probs,file=paste0("lda-gibbs_",18,"_topic-probs.csv"))
write.csv(phenomes_terms,file=paste0("lda-gibbs_",18,"_terms-to-topics.csv"))
```

visualise using heat-map

```{r}
heatmap(as.matrix(phenomes_probs),Colv=NA,scale="column")
```


